# Robust spatiotemporal mountain pine beetle example

```{r, echo = FALSE, cache=FALSE, message=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.height=8.5,
  fig.width=9
)
knitr::opts_knit$set(root.dir = "../../")
library(dplyr)
library(ggplot2)
library(glmmfields)
library(rstan)
library(viridis)
library(assertthat)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
# devtools::install_github("seananderson/ggsidekick")
theme_set(ggsidekick::theme_sleek())
```

Read in and set up the data.

```{r, echo=FALSE}
library(mapdata)
library(maps)
library(rgdal)
mpc <- ggplot2::map_data("worldHires", "Canada")
mps <- ggplot2::map_data("state")

mpc$group <- mpc$group + max(mps$group)
mp <- rbind(mpc,mps)

mp2 <- mp
coordinates(mp2) <- c("long", "lat")
proj <- "+proj=aea +lat_1=29.5 +lat_2=45.5 +lat_0=23 +lon_0=-96 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs +ellps=GRS80 +towgs84=0,0,0"
proj4string(mp2) <- CRS("+proj=longlat +datum=NAD83")
proj4string(mp2)

mp2 <- spTransform(mp2, CRS(proj))
proj4string(mp2)
```

```{r preparation, cache=FALSE, eval=TRUE}
id <- "mountain-pine-beetle-pnw-raster"
nbin <- 500

d <- readRDS(paste0("examples/beetles/",
  id, "-", "dataframe", "-", nbin, "x", nbin, ".rds"))

d_highres <- readRDS(paste0("examples/beetles/",
  id, "-", "dataframe", "-", nbin, "x", nbin, "highres.rds"))

# rescale
d$x <- d$x/1e5
d$y <- d$y/1e5

d <- dplyr::filter(d, cover > 0)
d_highres <- dplyr::filter(d_highres, cover > 0)

ggplot(d_highres, aes(x, y)) +
  # geom_polygon(data = mp2, aes(long, lat, group = group),
    # inherit.aes = FALSE, fill = "white", col = "grey40", lwd = 0.3) +
  geom_point(size = 0.1, alpha = 0.1) +
  facet_wrap(~year)
  # theme(panel.background = element_rect(fill="#e6e8ed")) +
  # coord_fixed(xlim = range(d_highres$x), ylim = range(d_highres$y))

ggplot(d, aes(x, y, fill = log(cover))) +
  # geom_polygon(data = mp2, aes(long/1e5, lat/1e5, group = group),
    # inherit.aes = FALSE, fill = "white", col = "grey40", lwd = 0.3) +
  geom_tile() +
  facet_wrap(~year) +
  scale_fill_viridis(option = "D") +
  coord_fixed()
  # theme(panel.background = element_rect(fill="#e6e8ed")) +
  # coord_fixed(xlim = range(d$x), ylim = range(d$y))

ggplot(d, aes(x, y, fill = cover)) + geom_tile() +
  facet_wrap(~year) +
  scale_fill_viridis(option = "D") +
  coord_fixed()

d <- mutate(d, row_number = seq_len(nrow(d)))
set.seed(1)
d <- group_by(d, year) %>%
  mutate(hold_out = row_number %in%
      sample(row_number, size = 25L)) %>%
  ungroup()

saveRDS(d, file = "examples/beetles/mountain-pine-beetle-data.rds")

# or the intercept prior will be off:
assert_that(mean(log(d$cover+0.5)) > -5 & mean(log(d$cover+0.5)) < 5)
# make sure data volume is reasonable:
assert_that(nrow(d) > 500)
assert_that(nrow(d) < 5000)
```

Fit the model. First the multivariate t model.

```{r model, cache=FALSE, warning=FALSE, message=FALSE, results='hide'}
model_file <- paste0("examples/beetles/", id, "-",
  "mvt-lognormal", "-", nbin, "x", nbin, ".rda")
if (!file.exists(model_file)) {
  mvt <- glmmfields(cover ~ 1,
    data = dplyr::filter(d, !hold_out),
    time = "year", lon = "x", lat = "y",
    nknots = 20,
    family = lognormal(link = "log"),
    prior_gp_sigma = half_t(3, 0, 1),
    prior_gp_theta = half_t(3, 0, 2),
    prior_intercept = student_t(1000, 0, 10),
    prior_rw_sigma = half_t(3, 0, 2),
    prior_sigma = half_t(3, 0, 3),
    prior_phi = half_t(1000, 0, 0.5),
    estimate_ar = TRUE,
    estimate_df = TRUE,
    year_re = TRUE,
    chains = 4, iter = 3000, thin = 1,
    control = list(adapt_delta = 0.95, max_treedepth = 20),
    # covariance = "exponential",
    # gp_sigma_scaling_factor = 1,
    save_log_lik = TRUE,
    df_lower_bound = 2)
  save(mvt, file = model_file)
} else {
  load(model_file)
}
# print(mvt)
# shinystan::launch_shinystan(mvt$model)
```

Now the multivariate normal model.

```{r, warning=FALSE, message=FALSE, results='hide'}
model_file <- paste0("examples/beetles/", id, "-",
  "mvn-lognormal", "-", nbin, "x", nbin, ".rda")
if (!file.exists(model_file)) {
  mvn <- glmmfields(cover ~ 1,
    data = dplyr::filter(d, !hold_out),
    time = "year", lon = "x", lat = "y",
    nknots = 20,
    family = lognormal(link = "log"),
    prior_gp_sigma = half_t(3, 0, 1),
    prior_gp_theta = half_t(3, 0, 2),
    prior_intercept = student_t(1000, 0, 10),
    prior_rw_sigma = half_t(3, 0, 2),
    prior_sigma = half_t(3, 0, 3),
    prior_phi = half_t(1000, 0, 0.5),
    estimate_ar = TRUE,
    estimate_df = FALSE,
    year_re = TRUE,
    fixed_df_value = 1000,
    chains = 4, iter = 3000, thin = 1,
    control = list(adapt_delta = 0.95, max_treedepth = 20),
    save_log_lik = TRUE)
  save(mvn, file = model_file)
} else {
  load(model_file)
}
# print(mvn)
# shinystan::launch_shinystan(mvn$model)
```

Let's make sure both models converged:

```{r}
bt <- broom::tidyMCMC(mvt$model, ess = TRUE, rhat = TRUE)
bn <- broom::tidyMCMC(mvn$model, ess = TRUE, rhat = TRUE)
max(bt$rhat)
min(bt$ess)
assertthat::assert_that(min(bt$ess) > 100)
assertthat::assert_that(max(bt$rhat) < 1.05)
max(bn$rhat)
min(bn$ess)
assertthat::assert_that(min(bn$ess) > 100)
assertthat::assert_that(max(bn$rhat) < 1.05)
```

Plot the predictions and check the model. Although functions for most of these plots are contained in the package, we will extract the data and make the plots ourselves so they can be more customized.

```{r plots, cache=FALSE, eval=TRUE}
pred <- predict(mvt, interval = "confidence", conf_level = 0.95,
  newdata = d)
pred <- pred %>% mutate(x = d$x, y = d$y, observed = d$cover,
  year = d$year, residual = log(observed) - estimate)

g <- ggplot(pred, aes(x, y, fill = estimate)) +
  geom_tile() + facet_wrap(~year) +
  scale_fill_viridis(option = "D") +
  coord_fixed()
print(g)

g <- ggplot(dplyr::filter(pred),
  aes(x, y, fill = exp(estimate))) +
  geom_tile() + facet_wrap(~year) +
  scale_fill_viridis(option = "D") +
  coord_fixed()
print(g)

g <- ggplot(pred, aes(x, y, fill = residual)) +
  geom_tile() + facet_wrap(~year) +
  scale_fill_gradient2() +
  coord_fixed()
print(g)

ggplot(pred, aes(estimate, residual)) +
  geom_point(alpha = 0.5) + facet_wrap(~year) +
  geom_hline(yintercept = 0, lty = 2) +
  geom_smooth(method = "loess", se = FALSE, colour = "red")

ggplot(pred, aes(log(observed), estimate)) +
  geom_point(alpha = 0.2) +
  geom_pointrange(aes(ymin = conf_low, ymax = conf_high), alpha = 0.1) +
  coord_fixed() +
  geom_abline(intercept = 0, slope = 1, colour = "red")
```

Let's compare the multivariate t and multivariate normal predictions.

```{r}
pred_mvn <- predict(mvn, interval = "confidence", conf_level = 0.95,
  newdata = d)
pred_mvn <- pred_mvn %>% mutate(x = d$x, y = d$y, observed = d$cover,
  year = d$year, residual = log(observed) - estimate)

g <- ggplot(dplyr::filter(pred_mvn),
  aes(x, y, fill = exp(estimate))) +
  geom_tile() + facet_wrap(~year) +
  scale_fill_viridis(option = "D") +
  coord_fixed()
print(g)

g <- ggplot(pred_mvn, aes(x, y, fill = residual)) +
  geom_tile() + facet_wrap(~year) +
  scale_fill_gradient2() +
  coord_fixed()
print(g)

pred2 <- pred
pred2$mvn_estimate <- pred_mvn$estimate
pred2$mvt_estimate <- pred$estimate
g <- ggplot(pred2, aes(mvt_estimate, mvn_estimate)) +
  geom_point() + facet_wrap(~year) +
  coord_fixed() +
  geom_abline(intercept = 0, slope = 1, colour = "red")
print(g)

g <- ggplot(pred2, aes(mvt_estimate, mvn_estimate)) +
  aes(x, y, fill = exp(mvt_estimate) - exp(mvn_estimate)) +
  geom_tile() + facet_wrap(~year) +
  scale_fill_gradient2() +
  coord_fixed()
print(g)

```

```{r}
combined <- bind_rows(
  mutate(pred, model = "mvt"),
  mutate(pred_mvn, model = "mvn"))

combined <- mutate(combined, id = paste(x, y))

set.seed(1234)
ids <- sample(unique(combined$id), size = 10)
jitter <- 0.2
dplyr::filter(combined, id %in% ids) %>%
  mutate(id_numeric = as.numeric(as.factor(id))) %>%
  mutate(id_numeric = ifelse(model == "mvn", id_numeric - jitter,
    id_numeric + jitter)) %>%
  ggplot(aes(x = id_numeric, y = estimate, ymin = conf_low,
    ymax = conf_high, color = model)) +
  geom_pointrange(size = 0.25) +
  facet_wrap(~year) +
  coord_flip() +
  geom_point(aes(x = round(id_numeric), y = log(observed)),
    color = "black", pch = 21) +
  xlab("y") + ylab("Point ID")
```

What about the holdout data?

```{r}
combined$hold_out <- c(d$hold_out, d$hold_out)

dplyr::filter(combined, hold_out) %>%
  ggplot(aes(log(observed), estimate)) +
  geom_point(alpha = 0.2) +
  geom_pointrange(aes(ymin = conf_low, ymax = conf_high), alpha = 0.1) +
  coord_fixed() +
  geom_abline(intercept = 0, slope = 1, colour = "red") +
  facet_wrap(~model)
```

Let's look at the mean squared error between the 2 models for the holdout data:

```{r}
mse <- dplyr::filter(combined, hold_out) %>%
  group_by(model, year) %>%
  summarize(mse = round(mean(residual^2), 6))
ggplot(mse, aes(year, mse, color = model)) +
  geom_line()

mse <- dplyr::filter(combined, hold_out) %>%
  group_by(model) %>%
  summarize(mse = round(mean(residual^2), 2))
print(mse)
```

What about the mean squared error for all of the data?

```{r}
mse <- dplyr::filter(combined) %>%
  group_by(model, year) %>%
  summarize(mse = round(mean(residual^2), 6))
ggplot(mse, aes(year, mse, color = model)) +
  geom_line()

mse <- dplyr::filter(combined) %>%
  group_by(model) %>%
  summarize(mse = round(mean(residual^2), 2))
print(mse)
```

Let's look at the observed and predicted values with confidence intervals for 10 of the holdout data points. In the simulations we've shown that the predictions in years with moderately large spikes can get inflated if we use the MVN model and the confidence intervals inflate to obtain the same correct coverage.

```{r}
jitter <- 0.2
dplyr::filter(combined, hold_out) %>%
  group_by(year, model) %>%
  mutate(id_numeric = 1:n()) %>%
  dplyr::filter(id_numeric <= 10) %>%
  mutate(id_numeric = ifelse(model == "mvn", id_numeric - jitter,
    id_numeric + jitter)) %>%
  ggplot(aes(x = id_numeric, y = estimate, ymin = conf_low,
    ymax = conf_high, color = model)) +
  geom_pointrange(size = 0.25) +
  facet_wrap(~year, scales = "free_x") +
  coord_flip() +
  geom_point(aes(x = round(id_numeric), y = log(observed)),
    color = "black", pch = 21) +
  xlab("y") + ylab("Point ID")
```

What about the width of the confidence intervals?

```{r}
cis <- mutate(combined, conf_width = conf_high - conf_low) %>%
  group_by(year, x, y) %>%
  summarize(conf_width_ratio = conf_width[model=="mvn"]/conf_width[model=="mvt"],
    hold_out = unique(hold_out)) %>%
  ungroup() %>%
  group_by(year) %>%
  mutate(median_ratio = median(conf_width_ratio))

dplyr::filter(cis) %>%
  ggplot(aes(conf_width_ratio)) +
  geom_histogram() +
  geom_vline(aes(xintercept = median(conf_width_ratio)))

dplyr::filter(cis) %>%
  ggplot(aes(conf_width_ratio)) +
  geom_histogram() +
  facet_wrap(~year) +
  geom_vline(xintercept = 1) +
  geom_vline(aes(xintercept = median_ratio), color = "red") +
  xlab("Ratio of prediction confidence intervals (MVN/MVT)")

median(cis$conf_width_ratio)
```

How are the model parameters themselves different?

```{r}
pars <- c("gp_sigma", "gp_theta", "sigma[1]")

bt <- broom::tidyMCMC(mvt$model, estimate.method = "median", conf.int = TRUE)
bn <- broom::tidyMCMC(mvn$model, estimate.method = "median", conf.int = TRUE)
bt$model <- "mvt"
bn$model <- "mvn"
b <- bind_rows(bt, bn)

dplyr::filter(b, term %in% pars) %>%
  ggplot(aes(term, estimate, ymin = conf.low,
    ymax = conf.high, color = model)) +
  geom_pointrange(position = position_dodge(width = 0.2)) +
  coord_flip()

bayesplot::mcmc_areas(as.matrix(mvt$model), pars = "df[1]")
```

Let's look at the coverage of the credible intervals between the 2 models:

```{r}
pred_mvt <- predict(mvt, interval = "prediction", conf_level = 0.95,
  newdata = d, type = "response")
pred_mvt <- pred_mvt %>% mutate(x = d$x, y = d$y, observed = d$cover,
  year = d$year, hold_out = d$hold_out)

pred_mvn <- predict(mvn, interval = "prediction", conf_level = 0.95,
  newdata = d, type = "response")
pred_mvn <- pred_mvn %>% mutate(x = d$x, y = d$y, observed = d$cover,
  year = d$year, hold_out = d$hold_out)

combined <- bind_rows(
  mutate(pred_mvt, model = "mvt"),
  mutate(pred_mvn, model = "mvn"))
```

Coverage for all the data:

```{r}
coverage <- dplyr::filter(combined) %>%
  group_by(model, year) %>%
  mutate(contained = observed < conf_high & observed > conf_low) %>%
  ungroup() %>%
  group_by(model) %>%
  summarize(coverage = mean(contained)) %>%
  print()
```

Coverage for just the held-out data:

```{r}
coverage <- dplyr::filter(combined, hold_out) %>%
  group_by(model, year) %>%
  mutate(contained = observed < conf_high & observed > conf_low) %>%
  ungroup() %>%
  group_by(model) %>%
  summarize(coverage = mean(contained)) %>%
  print()
```
