---
output:
  pdf_document: default
  html_document: default
---

# Example of Fish toxicity (Hg concentrations)

Data taken from EMMMA, https://emmma.usgs.gov/datasets.aspx. Samples from Wisconsin used 1985-1994. Different spp accumulate samples slightly differently, so we modeled Hg concentration for a single species (Black Crappie). We restricted the analysis to (1) samples that contained single fish sampled, and (2) sites with 3 or more samples collected. 

```{r packages, echo=FALSE, silent=FALSE, warning=FALSE}
library(sp)
library(rgdal)
library(ggplot2)
library(rrfields)
library(rstan)
library(dplyr)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())
```

## Load data

```{r load}
d = read.csv("examples/National Fish Data/National_Fish2_WI.csv")
d = d[which(d$CommonName=="Black Crappie" & d$NumFish == 1 & d$SampleType == "SKIN ON FILLET "),]

d$lon = d$LongDD
d$lat = d$LatDD

d = filter(d, Year %in% seq(1985, 1994)) %>%
  filter(lat != -999 & lon != -999) 
d = group_by(d, StationID) %>% 
  mutate(n = n()) %>% 
  filter(n >= 3) %>% ungroup
d$StationID = as.numeric(as.factor(d$StationID))

d$ID = seq(1,nrow(d))
coordinates(d) = c("lon","lat")
proj4string(d) <- CRS("+proj=longlat +datum=WGS84")  ## for example
d = as.data.frame(spTransform(d, CRS(paste("+proj=utm +zone=11"," ellps=WGS84",sep=''))))
d$lat = d$lat/1000000
d$lon = d$lon/1000000
d$Year = d$Year - min(d$Year) + 1

holdout = sample(1:nrow(d), size=round(nrow(d)*0.1,0), replace=F)
mvt_gamma = rrfield(HgResult ~ Length, 
  data = d[-holdout,], 
  time = "Year", lon = "lon", lat = "lat", station = "StationID", 
  nknots = 15, 
  obs_error = "gamma", 
  prior_gp_sigma = half_t(100, 0, 1), 
  prior_gp_scale = half_t(100, 0, 1), 
  prior_intercept = student_t(100, 0, 1), 
  prior_beta = student_t(100, 0, 1), 
  prior_sigma = half_t(100, 0, 1), 
  estimate_df = TRUE, 
  chains = 4, iter = 5000)

mvt_gamma

mvn_gamma = rrfield(HgResult ~ Length, 
  data = d[-holdout], 
  time = "Year", lon = "lon", lat = "lat", station = "StationID", 
  nknots = 15, 
  obs_error = "gamma", 
  prior_gp_sigma = half_t(100, 0, 1), 
  prior_gp_scale = half_t(100, 0, 1), 
  prior_intercept = student_t(100, 0, 1), 
  prior_beta = student_t(100, 0, 1), 
  prior_sigma = half_t(100, 0, 1), 
  estimate_df = FALSE, 
  chains = 4, iter = 5000)

#plot of data
#ggplot(d, aes(lon, lat,color=log(HgResult))) + 
# geom_point(alpha=1,size=1) + facet_wrap(~Year)

```

## Calculate prediction intervals

```{r intervals}
# Plot the predictions from each object to the fitted data
mvt_pred = predict(mvt_gamma, mcmc_draws = 1000,
  conf_level = 0.95, interval = "prediction")
mvn_pred = predict(mvn_gamma, mcmc_draws = 1000,
  conf_level = 0.95, interval = "prediction")

mvt_pred = predict(mvt_gamma, newdata = holdout, mcmc_draws = 1000,
  conf_level = 0.95, interval = "prediction")
mvn_pred = predict(mvn_gamma, newdata = holdout, mcmc_draws = 1000,
  conf_level = 0.95, interval = "prediction")
```


```{r, echo=FALSE, fig.pos="placeHere", fig.cap = "Estimated posterior degrees of freedom for the multivariate-t distribution"}
df=data.frame("x"=extract(mvt_gamma$model)[["df"]])
ggplot(data=df, aes(x=x)) + geom_histogram(alpha=0.5) + xlab("MVT degrees of freedom")
```


```{r, echo=FALSE, fig.pos="placeHere", fig.cap = "Estimated predicted vs observed (blue = mvt)"}
plot(log(mvn_pred$estimate), log(holdout[,"sum"]), col="red",
  pch = 16, cex=0.6, xlab = "Prediction", ylab = "Observed")
points(log(mvt_pred$estimate), log(holdout[,"sum"]), cex=0.8, col="blue")
abline(0,1)
```

```{r, echo=FALSE, fig.pos="placeHere", fig.cap = "Ratio of MVN CI widths : MVT CI widtsh"}

ratio = log((mvn_pred$conf_high - mvn_pred$conf_low) / (mvt_pred$conf_high - mvt_pred$conf_low))

hist(ratio, 40, col="grey", xlab = paste0("MVN CI widths : MVT CI widths"), main="")
```


```{r, echo=FALSE, fig.pos="placeHere", fig.cap = "Estimated year effects"}

mvt_year=extract(mvt_gamma$model)[["yearEffects"]]
mvt_year = data.frame("year"=seq(min(d$year),max(d$year)), 
  "mean"=apply(mvt_year,2,mean), "lower"=apply(mvt_year,2,quantile,0.025),
  "upper"=apply(mvt_year,2,quantile,0.975),"model"="mvt")
mvn_year=extract(mvn_gamma$model)[["yearEffects"]]
mvn_year = data.frame("year"=seq(min(d$year),max(d$year)), 
  "mean"=apply(mvn_year,2,mean), "lower"=apply(mvn_year,2,quantile,0.025),
  "upper"=apply(mvn_year,2,quantile,0.975),"model"="mvn")
mvt_year = rbind(mvt_year,mvn_year)

ggplot(data=mvt_year) + geom_ribbon(aes(x=year,ymin=lower,ymax=upper,
  group=model,fill=model),alpha=0.2) + 
  xlab("Year") + ylab("Estimated year effect")
```

```{r, echo=FALSE,fig.pos="placeHere", fig.cap = "Histogram of raw data and estimated values"}

par(mfrow = c(3,1), mai=c(0.5,0.5,0.1,0.1))
hist(holdout[,"sum"], 40, col="grey", main="Raw", xlab="", ylab="")
hist(mvt_pred$estimate, 40, col="grey", main="MVT", xlab="", ylab="")
hist(mvn_pred$estimate, 40, col="grey", main="MVN", xlab="", ylab="")
```
